---
title: "nlcor: Nonlinear Correlation"
author:
- name: "Chitta Ranjan"
  affiliation: "ProcessMiner, Inc."
  email: "cranjan@processminer.com"
- name: "Vahab Najari"
  affiliation: "ProcessMiner, Inc."
  email: "vnajari@processminer.com"
package: "nlcor"
date: "`r Sys.Date()`"
# output:
#   rmarkdown::html_vignette:
#     fig_caption: yes
output:
  md_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Purpose 
Estimate nonlinear correlations using `nlcor`. Yields a correlation 
estimate between 0 and 1, and the adjusted p value. The p value indicates if
the estimated correlation is statistically significant.

## Description
Correlations are commonly used in various data mining applications. 
Typically linear correlations are estimated. However, in many instances the data
has a nonlinear correlation but little to no linear correlation. If we are
performing data exploration using automated techniques on several variables,
such nonlinearly correlated variables can easily be overlooked.

Nonlinear correlations are quite common in real data. Due to this, nonlinear
models, such as SVM, are employed regression, classification, etc.
However, there are not many approaches to estimate nonlinear correlations for
application in data exploration, variable selection, and other area.

In this package, we provide an implementation of nonlinear correlation 
estimation using an adaptive local linear correlation computation. The function
`nlcor` returns the nonlinear correlation estimate, the corresponding 
adjusted p value, and an optional plot visualizing the nonlinear relationships. 

The correlation estimate will be between 0 and 1. The higher the value the more 
nonlinear correlation. Unlike linear correlations, a negative value is not 
valid here. Due to multiple local correlation computations, the net p value
of the correlation estimate is adjusted (to avoid false positives). The plot
visualizes the local linear correlations.

In following, we will show its usage with few examples. In the given examples,
the linear correlations between `x` and `y` is small, however, there is a 
visible nonlinear correlation between them. This package contains the data
for these examples and can be used for testing the package.

### Example 1.

A data with cyclic nonlinear correlation.


```{r, Figure-1.1, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
plot(x1, y1)
```

The linear correlation of the data is,
```{r}
cor(x1, y1)
```

As expected, the correlation is close to zero. We estimate the nonlinear
correlation using `nlcor`.

```{r, Figure-1.2, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
c <- nlcor(x1, y1, plt = T)
c$cor.estimate
c$adjusted.p.value
print(c$cor.plot)
```

The plot shows the piecewise linear correlations present in the data.

### Example 2.

A data with non-uniform piecewise linear correlations.

```{r, Figure-2.1, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
plot(x2, y2)
```

The linear correlation of the data is,
```{r}
cor(x2, y2)
```

The linear correlation is quite high in this data. However, there is 
significant and higher nonlinear correlation present in the data. 
This data is a emulates the scenario where the correlation changes its
direction after a point. Sometimes that change point is in the middle causing
the linear correlation to be close to zero. Here we show example when the 
change point is off center to show that the implementation works in 
non-uniform cases.

We estimate the nonlinear correlation using `nlcor`.

```{r, Figure-2.2, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
c <- nlcor(x2, y2, plt = T)
c$cor.estimate
c$adjusted.p.value
print(c$cor.plot)
```

It is visible from the plot that `nlcor` could estimate the piecewise
correlations in a non-uniform scenario. Also, the nonlinear correlation comes
out to be higher than the linear correlation.

### Example 3.
A data with higher and multiple frequency variations.

```{r, Figure-3.1, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
plot(x3, y3)
```

The linear correlation of the data is,
```{r}
cor(x3, y3)
```

The linear correlation is expectedly small, albeit not close to zero due to 
some linearity.

Here we show we can refine the granularity of the correlation computation.

Under default settings, the output of `nlcor` will be,

```{r, Figure-3.2, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
c <- nlcor(x3, y3, plt = T)
c$cor.estimate
c$adjusted.p.value
print(c$cor.plot)
```

As can be seen in the figure, `nlcor` overlooked some of the local relationships.
We can refine the correlation estimation by changing the `refine` parameter. 
The default value of `refine` is set as 0.5. It can be set as any 
value between `0` and `1`. A higher value enforces higher refinement. However,
higher refinement adversaly affects the p value. Meaning, the resultant 
correlation estimate maybe statistically insignificant (similar to overfitting).
Therefore, it is recommended to avoid over refinement.

In this data, we rerun the correlation estimation with `refine = 0.9`.

```{r, Figure-3.3, fig.show='hold', fig.align = 'center', fig.height = 4, fig.width = 6}
c <- nlcor(x3, y3, refine = 0.9, plt = T)
c$cor.estimate
c$adjusted.p.value
print(c$cor.plot)
```

As can be seen in the figure, `nlcor` could identify the granular piecewise
correlations. In this data, the p value still remains extremely small---the
correlation is _statistically significant_.

